---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# beastr

<!-- badges: start -->
<!-- badges: end -->

**B**etter **E**nvironment/**A**nimal **S**ensor **T**elemetry **R**epository

"Better" as in "better than nothing". There are many other solutions that are more
developed than this, most notably [Movebank](https://www.movebank.org/)

The goal of beastr is to provide a framework for storing, accessing, and processing
wildlife telemetry data. 

## Installation

You can install the package using the [devtools](https://devtools.r-lib.org/) package

``` r
devtools::install_github('rgzn/beastr')
```

## Example

Build your database:

```{r example}
library(beastr)
## basic example code
```

View data linked to animals, rather than sensors:

```{r cars}
#summary(cars)
```

## Related Projects

  - ctmmweb [https://github.com/ctmm-initiative/ctmmweb]
  - collardb [https://github.com/kissmygritts/collardb]
  - amt [https://github.com/jmsigner/amt]
  - movebank [https://www.movebank.org/]
  
##  TODO
  - integrate with amt (use amt tools)
  - integrate with movebank (export/import)
  - add new readers as new devices are used
