---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# beastr

<!-- badges: start -->
<!-- badges: end -->

**B**etter **E**nvironment/**A**nimal **S**ensor **T**elemetry **R**epository

"Better" as in "better than nothing". There are many other solutions that are more
developed than this, most notably [Movebank](https://www.movebank.org/)

The goal of beastr is to provide a framework for storing, accessing, and processing
wildlife telemetry data. 

## Installation

You can install the package using the [devtools](https://devtools.r-lib.org/) package

```{r, eval = FALSE}
devtools::install_github('rgzn/beastr')
```

## Example

Build your database:

```{r example, eval=TRUE, warning=FALSE}
library(beastr, quietly = TRUE)
library(sf, quietly = TRUE)

# Use example source data
fix_file = system.file("lotek/33452.txt", package = "beastr")
device_file = system.file("devices/collars.csv", package = "beastr")
animal_file = system.file("animals/critters.csv", package = "beastr")
deploy_file = system.file("deployments/deployments.csv", package = "beastr")
myDB = paste0(tempdir(check = TRUE), "/", "example.gpkg")

# Build a database
build_database(fix_files = fix_file,
               device_files = device_file,
               animal_files = animal_file,
               deployment_files = deploy_file,
               dsn = myDB,
               tz = "US/Pacific")

# What layers are in there?
sf::st_layers(myDB)
```

View data linked to animals, rather than sensors:

```{r eval=TRUE, warning=FALSE}
library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)

points = sf::st_read(myDB, layer = "animal_fixes")

points %>% 
  ggplot2::ggplot() + 
  ggplot2::geom_sf(ggplot2::aes(fill = animal_id))

```

## Related Projects

  - ctmmweb [https://github.com/ctmm-initiative/ctmmweb]
  - collardb [https://github.com/kissmygritts/collardb]
  - amt [https://github.com/jmsigner/amt]
  - movebank [https://www.movebank.org/]
  
##  TODO
  - integrate with `{amt}` (use amt tools)
  - integrate with movebank (export/import)
  - add new readers as new devices are used
  - use `{golem}` for the shiny app stuff
