---
title: "2. Filtering Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Filtering Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(beastr)
```


## Motivation

_What is "bad" data?_

Before we analyze data, we need to make sure it's not bogus. With GPS telemetry
we get a lot of questionable data. 

All of the data recorded is useful, so we won't
remove it outright. For location data, the most common type of bad data is a 
missed fix. This is where the collar did not find enough satellites to obtain a
location. This data, however, is still useful. The inability to find satellites
is not random. It is associated terrain features. It is more likely to find 
signals on the top of a mountain than the bottom of a canyon. If we were to 
analyze an animal's habitat use without taking these missed fixes into account, 
it would be biased towards the tops of ridges and peaks. 

Beside large scale terrain features, this missed data can give us information 
about microhabitat and animal behavior.

Here's an example:

```{r fig.width=6, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

# example database:
dsn = system.file("db/telemetry.gpkg", package = "beastr")

# first make sure the example data is in the example database:

# Recursing a directory to find lotek fix files:
fix_files = fs::dir_ls(system.file("lotek/", package = "beastr"),
                       regexp = "[/\\]PinPoint[ 0-9-]*.txt$",
                       recurse = TRUE)

# Add those fixes to database:
append_database(dsn = dsn, 
                fix_files = fix_files)

# Get points:
get_animal_fixes(dsn) %>%
  ggplot(aes(x=temp_c)) +
  geom_density(aes(fill = fix_status ), binwidth = 1, alpha = 0.7) +
  ggtitle("Fix Success vs Temp")

```

In this plot, we can see that the missed fixes are highly associated with the 
collar temperature. This indicates that the animals are resting in spots with 
poor satellite view. (Higher temperatures indicate the sensor is reading the 
animal's body rather than the air.)

Given that caveat about bad data, missed fixes don't actually have explicity 
spatial coordinates associated with them. So we will remove them for mapping. 

```{r fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
library(mapview) 

get_animal_fixes(dsn) %>% 
  filter(fix_status == "Valid") %>% 
  mapview(zcol = "animal_id")
```

You may notice some outlying points. Obviously we have fixes with actual 
coordinates that still aren't good. How can we remove these?

## Filters

There are quite a few different measures of how bad a fix is. Some of these come 
with the data, and some we have to compute ourselves. The most reliable approach 
is to use multiple measures and filter the data based on thresholds. Picking these
thresholds is somewhat mystical. Let's take a look at some of them:

### Dilution of Precision

Dilution of Precision (DOP) measures are calculated by the GPS receiver based on
satellite precision. There are multiple variants. Common ones are horizontal and
vertical. THe closer together the satellites are in the sky, the les 

Lotek gives us horizontal dilution of precision (hdop).  Let's take a 
look at the distribution from our example dB:

```{r fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
get_animal_fixes(dsn) %>% 
  ggplot(aes(x = hdop)) +
  geom_histogram() +
  xlab("hdop, (log scale)") +
  scale_x_log10()
```


