---
title: "1. Introduction"
output:
  rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{1. Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE, message = FALSE}
library(beastr)
library(dplyr)
library(readr)
library(ggplot2)
```

## Getting Started
_Help I have no idea what to do_

This document goes over what data we need to assemble a database. It is likely 
that you have this data already, but it may not be in the correct formats. The
goal here is to give the user an idea of a principled way to organize their data.
Unfortunately, assembling, formatting, and organizing data are not steps that 
can be automated. 

The example files here are all part of the package, so they live 
in subdirectories of:
```{r, eval=FALSE}
# not evaluated because it illustrates a directory on author's machine:
system.file("", package = "beastr")
```
And the directory structure therein represents one (of many) possibly ways to 
structure your data directories.

If you run this document from the .Rmd source, all the code sections like the
above can be run on your computer. Some sections, like the above are specified 
not to automatically evaluate. 

If you would like more information about RMarkdown, or just R in general, please 
see [the vignette on learning R](LearningR.html).


## Motivation
_Why would I bother?_



Why does anyone need a database? What data goes in here? I will briefly try to 
answer those questions here. 

Wildlife telemetry data is collected from placing devices in the field or directly
on subject animals. These devices record data such as GPS fixes, activity logs, photos, 
etc. 

We want to store and access not only information recorded by these devices, but 
information intrinsic to the devices (serial #, weight, etc) and the animals 
themselves (species, age, zodiac sign, etc). To maintain data validity, we 
can use separate tables to store each type of data, and then link those tables
together for our analyses. This prevents us from having to do things like having 
to store all the information about Ricky the Weasel (eg. Favorite Food: Squirrel)
on each of the twenty thousand records from the accelerometer on his collar.

The principle of storing each piece of information only once is called 
_atomicity_ and is key in preventing data mismatches as records are added or 
changed. 

For more motivation, please see the book that inspired this design:
_Spatial Database for GPS Wildlife Tracking Data_
[https://doi.org/10.1007/978-3-319-03743-1]()

And the basic principles of database theory:
[ACID](https://en.wikipedia.org/wiki/ACID)

And lastly, while the database backend is admittedly tedious, it is essential
for processing data, as in these guides:

  - Telemetry Database Data Entry
  - Wildlife Telemetry Quality Control
  - Exploring Fisher Collar Data

## Data Requirements
_What am I even dealing with?_

To build a new database, you will need to assemble the basic info as flat text 
files. This is the hardest part of the whole process, as it cannot be scripted. 

__Telemetry Data:__

This guide deals with data from [Lotek](https://www.lotek.com/) telemetry
devices. The principles and architecture may be extended to other devices, but
some new data parsing code will be required. This assumes you have data processed
with Lotek software that can be saved in Lotek's text file format. 

These files are produced using Lotek's proprietary software. They look like 
tab-delimitted text files, but there are some crucial differences. Let's take a 
look at an example file:

```{r}
# Example Lotek Telemetry data:
lotek_file = system.file("lotek/PinPoint33452.txt", package = "beastr")
readr::read_lines(lotek_file, n_max = 10)
```

You may notice some small details that make this file slightly difficult. There
are empty fields, but there is no dedicated delimiter to separate them. The
date/time fields are not standard formats and no coordinate reference system is
specified. Also there is crucial information (Device ID) only specified in the 
filename. 
If we want to use this data, import it to R with the `read_lotek()`
function:

```{r}
lotek_data = read_lotek(lotek_file)
head(lotek_data)
```

Now we have tidy spatial data, and can do spatial manipulation with it.
```{r}
ggplot(lotek_data) + 
  geom_sf() +
  ggtitle("Some Points from a Lotek Collar")
```

Telemetry devices may have other sensors as well. An example of this is "activity"
data. These records are records from the on-board accelerometer aggregated into
a measure called [Overall Dynamic Body Acceleration (ODBA)](). These records
are recorded on a different schedule from the GPS fixes, so must be stored 
separately. Again, Lotek uses a file format that is *almost* a standard csv, but 
not quite, so we'll need to use the function `read_lotek_activity()` from this
package:

```{r}
# Starts as 2 columns, switches to 3, doesn't parse:
system.file("lotek/activity33452.csv", package = "beastr") %>% 
  readr::read_delim(delim = ",", show_col_types = FALSE, n_max = 6)

# Using beastr:
system.file("lotek/activity33452.csv", package = "beastr") %>% 
  read_lotek_activity()
```


_{TODO}_: _Add support for more input file formats. This can be done as it 
becomes necessary with new devices. To preserve data, records with different 
fields for different types of devices should be stored in different tables. The
crucial information is joined together into a single `VIEW` using the `deployments` table._ 

__Device Data:__

Beyond the data recorded by the telemetry devices, we also want to store 
information about the devices themselves. This includes things like the serial
number, manufacturer, attachment type, etc. This is data tied to a device that 
does not vary between the individual samples recorded by the device. 

This data does not automatically come from the device, so it needs to be 
manually assembled into one or more CSV/spreadsheets. 

Here's what an example looks like:

```{r}
device_file = system.file("devices/collars.csv", package = "beastr")
readr::read_csv(device_file) %>% head(4)
```

In this case, we only have one list of devices. But if there were others that 
we wanted to combine, we could search for them using `fs`:

```{r}
# Find all CSVs in the device directory:
device_dir = system.file("devices", package = "beastr")
fs::dir_ls(device_dir, regexp = "*.csv", recurse = TRUE)
```


__Animal Data:__

The whole reason we are doing this is to learn about animals. We need to store 
information about them! A lot of what we know about the animals does not come 
from a telemetry device, we need to compile it here. 

Again, this will be one ormore flat CSV/spreadsheets. 

Here's what an example looks like, this time we're reading multiple files in:
```{r}
animal_dir = system.file("animals", package = "beastr")
animal_files = fs::dir_ls(animal_dir, regexp = "*.csv", recurse = TRUE)
readr::read_csv(animal_files) 
```

__Deployment Data:__
Animal telemetry devices are deployed on animals, but some animals may have 
multiple devices, and devices often record data even when they are not deployed.
In order to properly tie the various recorded data to their animals, we need to 
store deployment info. The key fields of this data are: _device_id_, _animal_id_,
_in_service_, _out_service_. There are more fields, but those are the key to 
understanding the purpose of this data. 

Again, this will be one or more flat CSV/spreadsheets. 

Example:

```{r}
deploy_file = system.file("deployments/deployments.csv", package = "beastr")
readr::read_csv(deploy_file)
```

_{TODO}_: _Need to add a non-animal deployment table for environmental and 
ambient devices. This should be a spatial table._

_Some telemetry devices may record environmental data (such as weather stations)
or ambient data (like Acoustic Recording Units), and are not tied to a specific 
animal. [NOT YET IMPLEMENTED]_


## Building a Geopackage

Now that you understand the raw data formats, we can run the script in the 
README:

```{r}
# Use example source data
fix_file = system.file("lotek/PinPoint33452.txt", package = "beastr")
device_file = system.file("devices/collars.csv", package = "beastr")
animal_file = system.file("animals/critters.csv", package = "beastr")
deploy_file = system.file("deployments/deployments.csv", package = "beastr")
activity_file = system.file("lotek/activity33452.csv", package = "beastr")

# Create a path to a geopackage in a temporary directory:
myDB = paste0(tempdir(check = TRUE), "/", "example.gpkg")

# Build a database
build_database(fix_files = fix_file,
               device_files = device_file,
               animal_files = animal_file,
               deployment_files = deploy_file,
               dsn = myDB,
               tz = "US/Pacific")

# What layers are in there?
sf::st_layers(myDB)
```


You may notice I forgot to import the activity files. Let's do that now, as an
example of adding an extra layer to an existing database:

```{r message=FALSE, warning=FALSE}
activity_data = read_lotek_activity(activity_file)
append_layer(data = activity_data,
             dsn = myDB,
             layer = "activity")

# a new layer:
sf::st_layers(myDB)

```


